No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-09-24 paper/GEB Submission/manuscript_GEB.Rmd
Line edits

  Results: We found no overall temporal trends in functional diversity metrics. Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities. We found evidence of an increasing temporal trend in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.

@@

Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g. coral bleaching, @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate [@boënnec2024; @johnson2024].
Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].

Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and where the space occurs across the range of possible trait distributions (community weighted trait means).
Considered together, they describe the aspects of the trait space that mediate ecological process.

Broad-scale trends in functional diversity are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, they can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.

@@

Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes using 6 different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five biomes (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
The earliest sample was in 1923 and the most recent was in 2014.
While it is not possible with available data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level. However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics. For a full breakdown of studies and their characteristics, see the supplement. Our final dataset reflects many of the data biases that make global synthesis work challenging, including geographic bias, a bias away from areas currently under the greatest threat, and a bias towards shorter time series. We address these shortcomings and their potential impact on our results in the discussion.

@@

We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine changes in the distribution of each trait.
Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained. CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).

@@

We estimated general trends across bird and mammal communities for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, methods which deal well with the inherent imbalances in our data. We fit 18 individual *CWM* models, one for each trait included in the analysis.
All time series with data for a given trait were included in the corresponding *CWM* model.
We estimated study level trends using individual linear models.
For studies with more than one location we fit a random slope and intercept for location.
Study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
They represented 13 of 1350 study-level models fit for each metric.
For further details see the supplement.
Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality.

To test for category-specific trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies. We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

category-specific trends in turnover highlight the types of communities experiencing change in community composition. For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies. The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities. We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.

@@

The lack of trend could be the result of multiple possible scenarios. First, these may be communities resisting perturbations or simply not experiencing significant perturbations. Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation. Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window. This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago [@estes2011].
Third, these communities may be experiencing directional shifts undetectable by available data. For example, species-level trait data do not capture intraspecific shifts in the trait space, which can have significant implications for ecological processes.

>>>>

  Results: We found no overall temporal trends in functional diversity metrics. Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities. We found evidence of a temporal increase in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.

@@

Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g. coral bleaching, @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate [@boënnec2024; @johnson2024].
Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, `r colorize("short time windows biased toward the present day", "blue")` and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].

Functional diversity offers a powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and `r colorize("how the space maps to individual trait distributions", "blue")` (community weighted trait means).
Considered together, they describe the aspects of the trait space that mediate ecological process.

`r colorize("Trends in functional diversity across broad geographic extents", "blue")` are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, they can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.

@@

Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes using `r colorize("7", "blue")` different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five biomes (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
The earliest sample was in 1923 and the most recent was in 2014.
While it is not possible with available data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.
For a full breakdown of studies and their characteristics, see the supplement.
Our final dataset reflects many of the data biases that make global synthesis work challenging, including geographic bias, a bias away from areas currently under the greatest threat, and a bias towards shorter time series.
We address these shortcomings and their potential impact on our results in the discussion.

@@

We also calculated the community-weighted mean (*CWM*) of all continuous traits `r colorize("(diet, foraging strata, body mass)", "blue")` to examine changes in the distribution of each trait. Wholesale shifts in the `r colorize("observed", "blue")` trait space due to changes of trait means could occur even while the shape `r colorize("and dispersion", "blue")` of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained `r colorize("through time", "blue")`. CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).

@@

We estimated general trends across bird and mammal communities for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, methods which deal well with the inherent imbalances in our data.
`r colorize("We fit individual models for each of the 18 CWM's calculated for continuous traits or trait categories (see diet, foraging strata, and body mass traits in Table 1).", "blue")` All time series with data for a given trait were included in the corresponding *CWM* model. We estimated study level trends using individual linear models. For studies with more than one location we fit a random slope and intercept for location. Study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models. They represented 13 of 1350 study-level models fit for each metric. For further details see the supplement. Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality. `r colorize("The log transformation was applied to all metrics except the SES corrected functional diversity metrics, FEve, and CWM's for`invertebrate`consumption, ground foraging strata, and below water surface foraging strata.", "blue")`

'r colorize("We tested for category-specific trends in species and functional diversity within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

Category-specific trends in `r colorize("species", "blue")` turnover highlight the types of communities experiencing change in community composition.
For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.
We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.

@@

The lack of trend could be the result of multiple possible scenarios.
First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation.
Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred `r colorize("thousands", "blue")` of years ago [@estes2011].
Third, these communities may be experiencing directional shifts undetectable by available data.
For example, species-level trait data do not capture intraspecific shifts in the trait space, which can have significant implications for ecological processes.
